\texttt{ensmallen}: a flexible C++ library for efficient function optimization

### 2018-10-18 doc/mlsys_paper/paper.tex
Minor reference fix.
Use eps format for the chart (labels included).
Set of changes for tonight:
- inline discussion of optimizers for Table 3
- space things out where they were a bit crowded
- a couple wordsmithing changes for space
- lengthen discussion on code and learning curves
- use subcaption for side-by-side figures (I thought it would save a little
  vertical space)
- add missing references

        & Julia ({\tt Optim.jl}) \cite{julia}         & \CIRCLE & - & - & \CIRCLE & - & - & - \\

@@

%% TODO: moving the list of optimizers here would improve the flow 
%% TODO: and free up a bit of space

@@

%% CS: i've added \nolimits to save a bit of space
\vspace*{-1.1em}
\begin{equation}
\argmindown_\theta f(\theta) =   %% CS: for clarity
\argmindown_\theta \sum\nolimits_{i = 1}^n (y_i - x_i \theta)^2 =
\argmindown_\theta \| y - X \theta \|_F^2.
\end{equation}
\vspace*{-1.1em}

@@

\vspace*{-1.1em}
\begin{equation}
f'(\theta) = \sum\nolimits_{i = 1}^n -2 x_i (y_i - x_i \theta) = -2 X^T (y - X \theta).
\end{equation}
\vspace*{-1.1em}

@@

\vspace*{-0.5em}
\begin{adjustbox}{scale={0.95}{0.95}}
\begin{minipage}{1\textwidth}
\begin{minted}[fontsize=\small]{c++}
    double EvaluateWithGradient(const arma::mat& theta, arma::mat& gradient) {
      const arma::vec v = (y - X * theta); // Cache result
      gradient = -2 * X.t() * v;  // Store gradient in the provided matrix
      return arma::accu(v % v); // Take squared norm of v
    }
\end{minted}
\end{minipage}
\end{adjustbox}
\vspace*{-0.5em}

% This is up here so that it's on the right page...
\begin{table}[t]
\begin{center}
\begin{tabular}{lcccc}
\toprule
 & {\tt ensmallen} & {\tt scipy} & {\tt Optim.jl} & {\tt samin} \\
\midrule
% TODO: these are just single-run results from Marcus' laptop!  We need to do
% 10 and average.
default & {\bf 0.004s} & 1.044s & 0.021s & 2.920s \\
tuned & & 0.5s-ish? & & ?? \\ % TODO
\bottomrule
\end{tabular}
\end{center}
\caption{Runtimes for $100000$ iterations of simulated annealing with various
frameworks on the simple Rosenbrock function.  Julia code runs do not count
compilation time.  The {\it tuned} row indicates that the code was manually
modified for speed.}
\label{tab:rosenbrock_results}
\end{table}

Template metaprogramming techniques are automatically used to
detect which methods exist, and a wrapper class will use suitable mix-ins in
order to provide any `missing` functionality~\cite{smaragdakis2000mixin}.  For
instance, if \texttt{\small EvaluateWithGradient()} is not provided, a version will be
automatically generated that calls both \texttt{\small Gradient()} and \texttt{\small Evaluate()} in turn.
Similarly, if \texttt{\small Evaluate()} or \texttt{\small Gradient()} does not exist, then {\tt
EvaluateWithGradient()} is called, and the unnecessary part of the result will
be discarded.

@@

@@

However, both Python and Octave have routes for acceleration,
such as Numba~\cite{TODO}, MEX bindings and JIT compilation.
We hand-optimized the Rosenbrock implementation using Numba,
which required significant modification of the 
underlying \texttt{\small anneal.anneal()} function.
We ran the Octave implementation with JIT enabled.
These techniques did produce some speedup,
as shown in the second row of Table~\ref{tab:rosenbrock_results}.
For Octave, a MEX binding did not produce a noticeable difference.

Next, we consider the linear regression example described in Sec.~\ref{sec:linreg_example}.
For this task we use the first-order L-BFGS optimizer~\cite{zhu1997algorithm}.
We compare across the following:

\vspace*{-0.3em}
\begin{enumerate}[{~~~$\bullet$}]
\renewcommand{\itemsep}{-0.5ex}
  \item {\tt ensmallen}-1: only \texttt{\small EvaluateWithGradient()} implemented
  \item {\tt ensmallen}-2: both \texttt{\small Evaluate()} and \texttt{\small Gradient()} separately
implemented
  \item {\tt Calculus.jl}: uses the default {\tt Calculus.jl} (\href{https://github.com/JuliaMath/Calculus.jl}{\footnotesize github.com/JuliaMath/Calculus.jl}) for
computing the gradient with {\tt Optim.jl}
  \item {\tt ForwardDiff.jl}: uses {\tt ForwardDiff.jl}~\cite{RevelsLubinPapamarkou2016} for computing
the gradient with {\tt Optim.jl}
  \item {\tt Optim.jl}: implements the gradient manually
  \item {\tt scipy}: implements both objective and
gradient
  \item {\tt bfgsmin}: uses {\tt optim} package for GNU Octave
\end{enumerate}
\vspace*{-0.3em}

@@

\begin{table}
\centering
\begin{adjustbox}{scale={0.90}{0.90}}
\begin{tabular}{lccccc}
\toprule
{\em algorithm} & $d$: 100, $n$: 1k & $d$: 100, $n$: 10k & $d$: 100, $n$:
100k & $d$: 1k, $n$: 100k \\
\midrule
% TODO: this was only one trial on Ryan's desktop!
{\tt ensmallen}-1 & {\bf 0.001s} & {\bf 0.006s} & {\bf 0.139s} & {\bf 1.698s} \\
{\tt ensmallen}-2 & 0.002s & 0.008s & 0.204s & 2.497s \\
{\tt Calculus.jl} & 0.172s & 0.960s & 27.428s & 2535.507s \\
{\tt ForwardDiff.jl} & 0.553s & 1.162s & 6.665s & 778.250s \\
{\tt Optim.jl} & 0.007s & 0.025s & 0.390s & 4.107s \\
{\tt scipy.optimize} & 0.002s & 0.010s & 0.198s & 2.022s \\
{\tt bfgsmin} & 0.215s & 1.705s & 15.998s & ??? \\
\bottomrule
\end{tabular}
\end{adjustbox}
\vspace*{0.25ex}
\caption{\footnotesize
Runtimes for the linear regression function on various dataset sizes,
with $n$ indicating the number of samples,
and $d$ indicating the dimensionality of each sample.
All Julia runs do not count compilation time.}
\label{tab:lbfgs}
\vspace*{-1ex}
\end{table}

The results indicate that {\tt ensmallen} with \texttt{\small EvaluateWithGradient()}
is the fastest approach.
Furthermore, the use of \texttt{\small EvaluateWithGradient()} yields considerable
speedup of almost 1.5x over the {\tt ensmallen} implementation with both the
objective and gradient implemented separately.  In addition, although the
automatic differentiation support makes it easier for users to write their
code, it is clear that the code generated by the automatic differentiators
({\tt Calculus.jl} and {\tt ForwardDiff.jl})
is far from optimal, with runs that are slower by up to three orders of magnitude.

@@

Lastly, we demonstrate the easy pluggability in {\tt ensmallen}
for using various optimizers on the same task.
Using a version of \texttt{\small LinearRegressionFunction} from Sec.~\ref{sec:linreg_example}
adapted for separable differentiable optimizers,
we run six optimizers with default parameters in just 7 lines of code,
as shown in Fig.~\ref{fig:learning_curve}(a).
Applying these optimizers to the \texttt{\small YearPredictionMSD}
dataset from the UCI repository~\cite{ucimlrepository}
yields the learning curves shown in Fig.~\ref{fig:learning_curve}(b).
Any other optimizers for separable differentiable objective
functions can be dropped into place in the same manner.

% This facilitates the evaluation of various optimizers
% for user-defined objective functions.

\begin{figure}[h!]
\begin{minipage}[t]{1\textwidth}
\begin{adjustbox}{scale={0.90}{0.90}}
\begin{minipage}{0.48\textwidth}
\centering
\begin{minted}[fontsize=\footnotesize]{c++}
// X and y are data
LinearRegressionFunction lrf(X, y);

ens::StandardSGD<>().Optimize(lrf, sgdModel);
ens::Adam().Optimize(lrf, adamModel);
ens::AdaGrad().Optimize(lrf, adagradModel);
ens::SMORMS3().Optimize(lrf, smorms3Model);
ens::SPALeRASGD().Optimize(lrf, spaleraModel);
ens::RMSProp().Optimize(lrf, rmspropModel);
\end{minted}
\end{minipage}
\end{adjustbox}
\hfill
\begin{minipage}{0.51\textwidth}
  \centering
  \includegraphics[width=\textwidth,height=0.45\textwidth]{experiments/learning_curves_crop.png}
\end{minipage}
\end{minipage}
\begin{minipage}[t]{1\textwidth}
\begin{minipage}{0.48\textwidth}
\centering
(a)
\end{minipage}
\hfill
\begin{minipage}{0.51\textwidth}
\centering
(b)
\end{minipage}
\end{minipage}
\caption
  {
  \footnotesize
  (a): example code for using 
  six {\tt ensmallen} optimizers with default parameters
  to optimize a linear regression function;
  (b): corresponding learning curves
  on the \texttt{\small YearPredictionMSD} dataset~\cite{ucimlrepository}
  over 5 epochs of training.
  The optimizers can be tuned for better performance.
  }
\label{fig:learning_curve}
\vspace*{-1ex}
\end{figure}

@@

We overviewed {\tt ensmallen}, a flexible C++ library for function
optimization that provides an easy interface for the implementation and optimization
of user-defined objective functions.  Many types of functions can be optimized,
including separable and constrained functions.
The library provides many pre-built optimizers (including numerous variants
of SGD and Quasi-Newton optimizers).
The library internally exploits template metaprogramming
to maximise opportunities for efficiency gains,
as well as 
to make the implementation of objective functions easier
by automatically generating missing methods.

% {\tt ensmallen} is already in use as the optimization toolkit of the {\tt
% mlpack} machine learning library~\cite{mlpack2018}.

>>>>

        & Julia ({\tt Optim.jl}) \cite{mogensen2018optim}         & \CIRCLE & - & - & \CIRCLE & - & - & - \\

@@

@@

%% CS: i've added \nolimits to save a bit of space
\vspace*{-0.5em}
\begin{equation}
\argmindown_\theta f(\theta) =   %% CS: for clarity
\argmindown_\theta \sum\nolimits_{i = 1}^n (y_i - x_i \theta)^2 =
\argmindown_\theta \| y - X \theta \|_F^2.
\end{equation}
\vspace*{-0.5em}

@@

\vspace*{-0.5em}
\begin{equation}
f'(\theta) = \sum\nolimits_{i = 1}^n -2 x_i (y_i - x_i \theta) = -2 X^T (y - X \theta).
\end{equation}
\vspace*{-0.5em}

@@

%\vspace*{-0.5em}
\begin{adjustbox}{scale={0.95}{0.95}}
\begin{minipage}{1\textwidth}
\begin{minted}[fontsize=\small]{c++}
    double EvaluateWithGradient(const arma::mat& theta, arma::mat& gradient) {
      const arma::vec v = (y - X * theta); // Cache result
      gradient = -2 * X.t() * v;  // Store gradient in the provided matrix
      return arma::accu(v % v); // Take squared norm of v
    }
\end{minted}
\end{minipage}
\end{adjustbox}
%\vspace*{-0.5em}

Template metaprogramming techniques are automatically used to
detect which methods exist, and a wrapper class will use suitable mix-ins in
order to provide `missing` functionality~\cite{smaragdakis2000mixin}.  For
instance, if \texttt{\small EvaluateWithGradient()} is not provided, a version will be
automatically generated that calls both \texttt{\small Gradient()} and \texttt{\small Evaluate()} in turn.
Similarly, if \texttt{\small Evaluate()} or \texttt{\small Gradient()} does not exist, then {\tt
EvaluateWithGradient()} is called, and the unnecessary part of the result will
be discarded.

@@

\begin{table}[t]
\begin{center}
\begin{tabular}{lcccc}
\toprule
 & {\tt ensmallen} & {\tt scipy} & {\tt Optim.jl} & {\tt samin} \\
\midrule
% TODO: these are just single-run results from Marcus' laptop!  We need to do
% 10 and average.
default & {\bf 0.004s} & 1.044s & 0.021s & 2.920s \\
tuned & & 0.5s-ish? & & ?? \\ % TODO
\bottomrule
\end{tabular}
\end{center}
\caption{Runtimes for $100000$ iterations of simulated annealing with various
frameworks on the simple Rosenbrock function.  Julia code runs do not count
compilation time.  The {\it tuned} row indicates that the code was manually
modified for speed.}
\label{tab:rosenbrock_results}
\end{table}

@@

% Actually Octave's JIT is apparently some kind of prototype joke and it doesn't
% even compile anymore.  So MEX was the only way...
However, both Python and Octave have routes for acceleration,
such as Numba~\cite{lam2015numba}, MEX bindings and JIT compilation.
We hand-optimized the Rosenbrock implementation using Numba,
which required significant modification of the
underlying \texttt{\small anneal.anneal()} function.
These techniques did produce some speedup,
as shown in the second row of Table~\ref{tab:rosenbrock_results}.
For Octave, a MEX binding did not produce a noticeable difference.
We were unable to tune either \texttt{\small ensmallen} or \texttt{\small
Optim.jl} to get any speedup, suggesting that novice users will easily be able
to write efficient code in these cases.

Next, we consider the linear regression example described in
Sec.~\ref{sec:linreg_example}.  For this task we use the first-order L-BFGS
optimizer~\cite{zhu1997algorithm}.  Using the same four packages, we implement
the linear regression objective and gradient.  For \texttt{\small ensmallen} we
implement a version with only \texttt{\small EvaluateWithGradient()}:
\texttt{\small ensmallen}.  We also implement a version with both \texttt{\small
Evaluate()} and \texttt{\small Gradient()}: \texttt{\small ensmallen}-2.  We
also use automatic differentiation for Julia via the \texttt{\small
ForwardDiff.jl}~\cite{RevelsLubinPapamarkou2016} package and for Python via the
\texttt{\small Autograd}~\cite{maclaurin2015autograd}
package.  For GNU Octave we use the \texttt{\small bfgsmin()} function.

@@

\begin{table}
\centering
\begin{adjustbox}{scale={0.90}{0.90}}
\begin{tabular}{lccccc}
\toprule
{\em algorithm} & $d$: 100, $n$: 1k & $d$: 100, $n$: 10k & $d$: 100, $n$:
100k & $d$: 1k, $n$: 100k \\
\midrule
% TODO: this was only one trial on Ryan's desktop!
{\tt ensmallen}-1 & {\bf 0.001s} & {\bf 0.006s} & {\bf 0.139s} & {\bf 1.698s} \\
{\tt ensmallen}-2 & 0.002s & 0.008s & 0.204s & 2.497s \\
% Dropped for space and awful performance
%{\tt Calculus.jl} & 0.172s & 0.960s & 27.428s & 2535.507s \\
{\tt Optim.jl} & 0.007s & 0.025s & 0.390s & 4.107s \\
{\tt scipy} & 0.002s & 0.010s & 0.198s & 2.022s \\
{\tt bfgsmin} & 0.215s & 1.705s & 15.998s & ??? \\
% It's possible to tune ForwardDiff.jl a bit, but it doesn't give significant
% speedups to make it competitive and it really makes the code ugly.
{\tt ForwardDiff.jl} & 0.553s & 1.162s & 6.665s & 778.250s \\
{\tt autograd} & ??? & ??? & ??? & ??? \\
\bottomrule
\end{tabular}
\end{adjustbox}
\vspace*{0.25ex}
\caption{\footnotesize
Runtimes for the linear regression function on various dataset sizes,
with $n$ indicating the number of samples,
and $d$ indicating the dimensionality of each sample.
All Julia runs do not count compilation time.}
\label{tab:lbfgs}
\vspace*{-1ex}
\end{table}

The results indicate that \texttt{\small ensmallen} with \texttt{\small
EvaluateWithGradient()} is the fastest approach.
Furthermore, the use of \texttt{\small EvaluateWithGradient()} yields considerable
speedup of almost 1.5x over the \texttt{\small ensmallen} implementation with both the
objective and gradient implemented separately.  In addition, although the
automatic differentiation support makes it easier for users to write their
code, it is often true that the output of automatic differentiators is not as
efficient, especially with \texttt{\small ForwardDiff.jl}.  We expect this
effect to be more pronounced with increasingly complex objective functions.

@@

Lastly, we demonstrate the easy pluggability in {\tt ensmallen}
for using various optimizers on the same task.
Using a version of \texttt{\small LinearRegressionFunction} from Sec.~\ref{sec:linreg_example}
adapted for separable differentiable optimizers,
we run six optimizers with default parameters in just 7 lines of code,
as shown in Fig.~\ref{fig:learning_curve}(a).
Applying these optimizers to the \texttt{\small YearPredictionMSD}
dataset from the UCI repository~\cite{ucimlrepository}
yields the learning curves shown in Fig.~\ref{fig:learning_curve}(b).
Any other optimizers for separable differentiable objective
functions can be dropped into place in the same manner.
% Just because we have some extra space...
This facilitates the seamless evaluation of various optimizers
for user-defined objective functions.

\begin{figure}[b!]
\centering
\begin{subfigure}[b]{0.49\textwidth}
\begin{minted}[fontsize=\footnotesize]{c++}
// X and y are data
LinearRegressionFunction lrf(X, y);

using namespace ens;
StandardSGD<>().Optimize(lrf, sgdModel);
Adam().Optimize(lrf, adamModel);
AdaGrad().Optimize(lrf, adagradModel);
SMORMS3().Optimize(lrf, smorms3Model);
SPALeRASGD().Optimize(lrf, spaleraModel);
RMSProp().Optimize(lrf, rmspropModel);
\end{minted}
\caption{Code.}
\end{subfigure}
\begin{subfigure}[b]{0.49\textwidth}
  \centering
  % \includegraphics[width=\textwidth,height=0.45\textwidth]{experiments/learning_curves_crop.png}
  \includegraphics[width=\textwidth,height=0.6\textwidth]{experiments/learning_curves_crop.eps}
\caption{Learning curves.}
\end{subfigure}
\caption{Example usage of six \texttt{\small ensmallen} optimizers to optimize a
linear regression function on the \texttt{\small YearPredictionMSD}
dataset~\cite{ucimlrepository} for 5 epochs of training.  The optimizers can be
tuned for better performance.}
\label{fig:learning_curve}
\vspace*{-1ex}
\end{figure}

@@

We have described {\tt ensmallen}, a flexible C++ library for function
optimization that provides an easy interface for the implementation and optimization
of user-defined objective functions.  Many types of functions can be optimized,
including separable and constrained functions.
The library provides many pre-built optimizers (including numerous variants
of SGD and Quasi-Newton optimizers).
The library internally exploits template metaprogramming
to maximise opportunities for efficiency gains,
as well as 
to make the implementation of objective functions easier
by automatically generating missing methods.

% RC: I think it's really important to highlight ensmallen's usage (and
% genesis), although I can't find the right words to concisely and non-awkwardly
% say that we wrote ensmallen as part of mlpack originally.
\texttt{\small ensmallen} is already in use as the optimization toolkit of the
\texttt{\small mlpack} machine learning library~\cite{mlpack2018}.
